Formal Construction of Microcontroller Instruction Set Model Using B

### 2010-08-02 papers/Article2010/source/paper.tex
Ainda colocando conteúdo ... falta escrever muita coisa

\documentclass{llncs}
%\documentclass{article}

\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{float}
%\usepackage{supertabular} % used in B Symbols

\include{comandos}
\include{SymbolsB_AtelierB}

@@

\author{ David D\'{e}harbe\inst{1}, Val\'{e}rio Medeiros Jr\inst{2} }
\institute{Federal University of Rio Grande do Norte, Natal RN 59078-970, Brazil\inst{1} and\\  
Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal RN
59015-000, Brazil\inst{2}}

@@

\begin{abstract}
  [To redo]
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform. This work is a contribution towards the extension of the
  B method to handle developments up to assembly level code.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

\section{Introduction}
[To update]
The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

This paper is focused on the presentation of the Z80 model, including
elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short
introduction to the B method. Section~\ref{sec:models} presents the
elementary libraries and the modelling of some elements common to
microcontrollers.  Section~\ref{sec:z80} presents the B model of the
Z80 instruction set. Section ~\ref{sec:Proofs} provides some
information on the proof effort needed to analyze the presented
models. Related work is discussed in
Section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

\section{Introduction to the B Method}
\label{sec:B_method}

@@

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition between states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modelling stage, the B method requires that the user proves that, in a machine,
all the its initial states are valid, and that operations do not define
transitions from valid states to invalid states.

@@

\begin{figure}
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{1cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

\section{Model structure and basic components}
\label{sec:models}

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of the platform. Thus, the workspace is
composed of: a power library\footnote{The power library has the basics definitions to help the theorems
prover about simple calculus of power, especially power of two. It has constants, that represents all
results of power function needed, and the definition of power function.}, a hardware library, a types
library and a project for the specific platform, in this case the Z80. The corresponding dependency
diagram is depicted in Figure~\ref{fig:hardware-definition-graph}; information specific to each project is
presented in the following.

@@

\subsection{Bit Representation and Manipulation}

@@

\subsection{Representation and Manipulation of Bit Vectors}

@@

$
\begin{array}{l}
\mathit{bv\_not} \in \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_not} = \lambda v . (v \in \mathit{BIT\_VECTOR} \mid \quad \lambda i . (1 .. \mathit{bv\_size}(v)) \mid \mathit{bit\_not}(v(i))) \land \\
\mathit{bv\_and} \in \mathit{BIT\_VECTOR} \times \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_and} = \lambda v_1, v_2 . (v_1 \in \mathit{BIT\_VECTOR} \land v_2 \in \mathit{BIT\_VECTOR} \land \\
\quad \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \mid \lambda i . (1 .. \mathit{bv\_size}(v_1)) \mid
\mathit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$

@@

\subsection{Modelling Bytes and Bit Vectors of Length 16}

@@

\subsection{Bit Vector Arithmetics}

@@

\subsection{Basics Data Types}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
 $Type\ Name$ & UCHAR & SCHAR & USHORTINT & SSHORTINT  & BYTE & BV16 \\\hline
 $Range$ & 0..255 & -128..127 & 0..65.535  & -32.768..32.767  & -- & --\\ \hline
 $Physical\ Size $ & 1 byte & 1 byte & 2 bytes & 2 bytes &  1 bytes & 2 bytes \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

% We also created the following lemmas:
%
% $
% \begin{array}{l}
%  \forall (val) . (val \in \mathit{UCHAR} |
%  \mathit{byte\_uchar}(\mathit{uchar\_byte}(val)) = val) \land\\
%  \forall (by) . (by \in \mathit{BYTE} |
%  \mathit{uchar\_byte}(\mathit{byte\_uchar}(by)) = by)
% \end{array}
% $

Similarly, several other functions and lemmas were created for all other data types.

\section{Description of the Z80 B model }
\label{sec:z80}

The \textit{Z80} is a CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports 158
different instructions, including all the 78 from Intel 8080 microprocessor, and all of them were
specified. These instructions are classified into these categories: load and exchange; block transfer and
search; arithmetic and logical; rotate and shift; bit manipulation; jump, call and return; input/output;
and basic cpu control.

@@

% Chacteristics of microcontroller
 The internal registers contain 176 bits of reading/writing memory that are represented by
identifiers used as parameters in the instructions . It includes two sets of six general purpose
registers which may be used individually as 8-bits registers or as 16-bits register pairs.  The working registers
are represented by variable $\mathit{rgs8}$. The domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set
formed by identifiers of registers of 8 bits. These registers can be accessed in pairs, forming 16-bits,
resulting in another set of identifiers of 16-bits registers, named $\mathit{id\_reg16}$.

@@

% \subsection{Modelling Registers, Input and Output Ports and Instructions}
% 
% The Z80 has different types of registers and instructions. The CPU
% contains general-purpose registers ($\mathit{id\_reg\_8}$), a stack
% pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two index
% registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register
% ($\mathit{i\_}$), a refresh register ($\mathit{r\_}$), two bits
% ($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions,
% a pair of bits to define the interruption mode ($\mathit{im}$) and the
% input and output ports ($\mathit{i\_o\_ports}$). Below, part of the
% corresponding definitions are replicated from the $\INVARIANT$:
% 
% \begin{sloppypar}
% %\bf INVARIANT
% \hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$
% 
% \hspace*{0.10in}\it  sp  $\in$  \it BV16  $\land$  \it ix  $\in$  \it BV16  $\land$  \it iy  $\in$  \it BV16  $\land$
% 
% \hspace*{0.10in}\it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$ iff1  $\in$  \it BIT
% $\land$ \it iff2  $\in$  \it BIT  $\land$
% 
% \hspace*{0.10in}\it im \rm $\in$ \rm (\it BIT $\times$ \it BIT\rm )  $\land$ i\_o\_ports  $\in$  \it BYTE
% $\fun$  \it BYTE
% \end{sloppypar}
% 
% A simple example of instruction is a $\mathit{LD\_n\_A}$, as shown
% below.  Many times, to model an instruction is necessary to use the
% predefined functions, these help the construction of model. This
% instruction use the $\mathit{updateAddressMem}$ function from
% \textit{Memory} module and it receives an address memory and its new
% memory value. Finally it increments the program counter
% ($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).
% 
% \hspace*{0.00in}\bf LD\_n\_A \rm ( \it nn \rm ) \rm =
% 
% \hspace*{0.20in}\bf PRE \it nn $\in$ \it USHORT\hspace*{0.15in} %$\land$  \it nn $\in$\it DATA\_R\_ADR
% 
% \hspace*{0.20in}\bf THEN
% 
% \hspace*{0.20in}\bf updateAddressMem \rm ( \it ushort\_to\_bv16 \rm ( \it nn \rm ) \rm , \it rgs8 \rm ( \it a0 \rm )
% \rm )  $\para$
% 
% \hspace*{0.20in}\it pc \rm := \it instruction\_next \rm ( \it pc \rm )  $\para$  \it r\_ \rm := \it
% update\_refresh\_reg\rm (\it r\_\rm )
% 
% \hspace*{0.00in}\bf END\rm
% 
% The microcontroller model can specify security properties. For example, the
% last operation could have a restriction to write only in a defined region of memory.

\subsection{Modelling registers and input and output ports} [I need review the english]

\hspace*{0.10in}\it i\_  $\in$  \it BYTE  $\land$  \it r\_\hspace*{0.10in} $\in$  \it BYTE  $\land$

\hspace*{0.10in}\it im \rm : \rm (\it BIT $\times$ \it BIT\rm )  $\land$

\subsection{Flag register} [I need review the english]
Another important element is the ``z'' register ($\mathit{rgs8(z0)}$),  that is used as a flag register. This
register uses only six bits to represent the result status of each instruction. 	
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning:
\begin{description}
  \item[$\mathit{bv\_get(rgs8(z0),0)}$] - The Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),1)}$] - The Add/Subtract bit
  \item[$\mathit{bv\_get(rgs8(z0),2)}$] - The Parity or Overflow bit
  \item[$\mathit{bv\_get(rgs8(z0),4)}$] - The Half Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),6)}$] - The Zero bit
  \item[$\mathit{bv\_get(rgs8(z0),7)}$] - The Sign bit 
\end{description}

\begin{property}[Assuring the absence of Overflow]
 To assure that don¥t happen overflow the developer can add this expression ($\mathit{bv\_get(rgs8(z0),0)} \neq 1
 \land \mathit{bv\_get(rgs8(z0),2) \neq 1}$) in the invariant. By default, the overflow can happen, but
 many times it can be dangerous. Then, the developer may prohibit its use. This restriction can also become the more dificult to
 specify.
\end{property}

\subsection{Manipulation data functions from Z80} [I need to review the english] There are some specific functions
from Z80 to manipulate the data. The $\mathit{bv\_ireg\_plus\_d}$ is  used to indexed address. It received the
value of register ($\mathit{ix}$ or $\mathit{iy}$) and displacement to return the sum, the result is the memory address dislocated, see
its the definition.

@@

Another derived function is  $\mathit{bv\_9ireg\_plus\_d0}$ , this returns the value in the memory  address
returned by $\mathit{bv\_ireg\_plus\_d}$ function ands its definition is similar.

@@

\subsection{Program, stack and data memory}[I need review the english]

The Z80 uses a unique memory for storing program instructions, data stack and data work. The memory has 16-bits
addressing and each address represent a byte. Thus, the data from the memory module is very simple, as shown below,
but an additional care must be taken to preserve the consistency of memory.

@@

In general, the instructions can access all memory address, but it is very dangerous. For add more security, is
importante that the program instructions has the access limited by region. Thus, the designer can specify address
regions to restrict the access from instructions. The address regions can be especified how shown below.

$
\begin{array}{l}
\mathit{PROGRAM\_R\_ADR} = 0..16384 \land \mathit{DATA\_R\_ADR} = 16385..49151 \land \\
\mathit{STACK\_R\_ADR} = 49152..65535
\end{array}
$

\begin{property}[Assuring the absence of overlapping of address regions]
 To assure that address regions are well defined, then the designer must to verify the below expression.
\end{property}

@@

\begin{property}[Preserving the consistency of the memory]
In general, the access to some address regions is dangerous. Then, in each instruction has a specific
pre-condition that verify if the new address memory, that will be updated, is member of its region. For example,
the $\mathit{PUSH}$ program instruction allow write only in the region of stack ($\mathit{STACK\_R\_ADR}$). 
%Thus,if the user try to write in incorrect region then the called of operation cannot be proved.
\end{property}

@@

There are many functions defined in the module \textit{ALU}. In general, these functions take basic definitions to build
new specific functions. The function $\mathit{half8UCHAR}$ is used to get the half part of $\mathit{UCHAR}$ value.
It is importante to know the half carry and it is used in the $\mathit{add8UCHAR}$.

@@

A similar function to subtract operation is $\mathit{substract8UCHAR}$. There are the same functions for the
$\mathit{SCHAR}$ type, they are respectively $\mathit{add8SCHAR}$ and $\mathit{substract8SCHAR}$, all these
functions are of 8 bits ($\mathit{BYTE}$) and defined similarly. In the same way, the arithmetic functions for 16
bits ($\mathit{BV16}$) are defined.

Other more simply functions also are typed and explained below:

\begin{itemize}
  \item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and returns its decrement. There is a
  similar function named $\mathit{dec}$.

  \item \it instruction\_next  $\in$  USHORT  $\fun$  USHORT \rm - It receives the pc actual
  value and return its increment.

  \item \it is\_negative  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
  byte is zero, otherwise returns 0.

  \item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.

@@

Every logic functions are defined in the \textit{BYTE} and \textit{BV16} module are included in the \textit{ALU}
module. Then, it was not necessary rewrite these functions.

\subsection{Modelling the instructions}[I need review the english]

Each instruction is represented by a B operation in the module Z80. The main module (\textit{Z80}) has two
types of operations, a type represents the external actions and the other represents the instructions of
microcontrollers. The external actions are shown in the section ~\ref{sec:externalactions}.  A simple
example from instruction is a $\mathit{LD\_(nn)\_A}$ shown below. Many times to model a instruction is
necessary to use the pre-defined functions, this facilitates the construction of model. By default, all
parameters from operations are either predefined elements in the model or integers values in the decimal
representation. This instruction use the $\mathit{updateAddressMem}$ function from \textit{Memory} module
and it receives a address memory and its new memory value. After, it increments the program counter
($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).

@@

\subsection{Modelling the Input/Output Instructions}

@@

The $\mathit{IN\_r(C)}$ \footnote{The tools B don't allow use parenteses in identifiers, then the characteres
\{``('',``)''\} are replaced respectively by \{``9'',``0''\} in the real specification.} instruction is represented
by the follow B operation. It is from I/O group, then it reveives a identifier of register
$\mathit{r}$ and, in this place, it stores the value of $\mathit{C}$ port address. Besides, it increments the
program counter and updates the flag registers.

@@

\subsection{Modelling the external action}
\label{sec:externalactions}

>>>>

%\documentclass{llncs}
\documentclass[11pt]{article} % use larger type;

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{float}
%\usepackage{supertabular} % used in B Symbols
\geometry{a4paper} % or letterpaper (US) or a5paper or....

\include{comandos}
\include{SymbolsB_AtelierB} 
% Allow use multiples footnote references 
\newcommand{\footnoteremember}[2]{
  \footnote{#2}
  \newcounter{#1}
  \setcounter{#1}{\value{footnote}}
}
\newcommand{\footnoterecall}[1]{
  \footnotemark[\value{#1}]
}

@@

\author{Val\'{e}rio Medeiros Jr%\inst{2}
, David D\'{e}harbe%\inst{1}
}

 %\institute{ \inst{1}Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal RN 59015-000, Brazil % } 
 %  \and %\institute{
 %  \inst{2}Federal University of Rio Grande do Norte, Natal RN 59078-970,Brazil}

@@

\begin{abstract}
  [TODO:To redo]
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform. This work is a contribution towards the extension of the
  B method to handle developments up to assembly level code.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

\section{Introduction}
[TODO:To update] 
The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

This paper is focused on the presentation of modelling of assembly instruction
set, including elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short introduction
to the B method. Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. Section ~\ref{sec:Verification}
provides some information on the proof effort needed to analyze the presented
models. Related work is discussed in Section~\ref{sec:relatedworks}. Finally, the
last section is devoted to the conclusions.

\subsection{Problems in formalizing instruction set}
[I need review the text and move this secion to other place]
%a) Explicar como sheo formalizados os conjuntos de instruheheo nos
%manuais dos construtores (Consultar os dois manuais)

In general, the manuals of microcontrollers can be more appropriate for
developers in assembly language. Many times, each instruction of microcontrollers
are shown in its official manual using textual description, math partially
description of instruction, examples, encoding, number of cycles and others
informations. But the explantion about the action of instruction do not use
usually the better way to describe the instruction for users and developers. For example,
if the user want to know the action of instruction and its effect on the flag
register then he needs search this information in different pages of manual.
Besides, the writer of manual usually describes important facts in not formal
notation and default. So this description can add some ambiguities or until
mistakes.

% b) The case of Z80
The official manual of the microcontroller Z80 \cite{Z80_manual} has several
problems that were identified over time~\cite{UndocumentedZ80}. The Z80 is a
microcontroller that  has been tested and used for many years. This intensive use
and test facilitated to find several errors in the description of the
instructions in its manual. Several of these problems have been cataloged in a
technical report \cite{UndocumentedZ80}. Some problems are inaccurate
information, for example the description of action of instruction on the flag
register, partial informations and distributed on different pages of manual.

Some erros was identified during the development of Z80 model.
An example is
%\begin{itemize}
% \item Some input and output instructionsThe
% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some official documentation says they leave it unaffected, important!) and the
% NF flag is not always set but may also be reset (see 4.3 for exact operation)
% \cite{UndocumentedZ80}.
% Este erro he descrito na phegina (274 do documento e 294 do pdf) 
% do manual oficial do Z80 (Z80_CPU_adotado.pdf)
% Citar apenas esse exemplo abaixo! 
%   \item
when a determined interruption (NMI) is accepted, the bit IFF1 is not
copied to bit IFF2. % Este erro he descrito na phegina (23 do documento e 41do pdf) % do manual oficial do Z80 (Z80_CPU_adotado.pdf)
Only IFF1 is reset. This problem was identified, because the description of
interruption on the official manual is big and complex, so to help the
construction of Z80 B model, others references  \cite{UndocumentedZ80,Simulator_z80} 
also were consulted.

%\end{itemize}

An interesting solution is to specify formally the assembly instruction set,
this creates a pattern of representation and validates properties of
instructions. The formal model also restricts the definitions to correct typing and use only
expressions well-defined.
% -> Solucao interessante he especificar  formalmente o conjunto
% de instruhehehees, isto cria um padrao formal de representaheheheo e valida propriedades
% das instrucoes. 
Furthermore, the developer can use with B method different levels of
abstraction and add more specific details in refined models.

% allow to verify properties on the model.
\section{Introduction to the B Method}
\label{sec:B_method}

@@

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition between states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modeling stage, the B method requires that the user proves that, in a machine,
all the its initial states are valid, and that operations do not define
transitions from valid states to invalid states.

@@

\begin{figure}[h]
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT\\ 
      \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{0cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

\section{Model structure and basic components}
\label{sec:models}
%Em cada subsecaoo descrever os principais lemas que foram provados.
%Sugestheo de David:
% a) structure
% b) formalization of bit-level concepts
% c) formalization of bit vectors
% d) formalization of byte and word-level concepts
% e) formalization of integer ranges

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped into three separate development projects
and are available as libraries. A fourth project is devoted to the higher-level aspects of the platform. Thus the workspace is
composed of: a power library\footnote{The power library has the basics definitions to help the theorems
prover about simple calculus of power, especially power of two. It has constants, that represents all
results of power function needed, and the definition of power function.}, a hardware library, a types
library and a project for the specific platform, in this case the Z80. The corresponding dependency
diagram is depicted in Figure~\ref{fig:hardware-definition-graph}; information specific to each project is
presented in the following.

@@

\subsection{Bit representation and manipulation}

@@

\subsection{Representation and manipulation of bit vectors}

@@

$
\begin{array}{l}
\mathit{bv\_not} \in \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_not} = \lambda v . (v \in \mathit{BIT\_VECTOR} \mid \quad \lambda i . (1 .. \mathit{bv\_size}(v)) \mid \mathit{bit\_not}(v(i))) \land \\
\mathit{bv\_and} \in \mathit{BIT\_VECTOR} \times \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_and} = \lambda v_1, v_2 . (v_1 \in \mathit{BIT\_VECTOR} \land v_2 \in \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \mid \lambda i . (1 .. \mathit{bv\_size}(v_1)) \mid
\mathit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$

@@

\subsection{Modelling bytes and bit vectors of length 16}

@@

\subsection{Bit vector arithmetics}

@@

\subsection{Basics data types}

@@

% \begin{center}
% \begin{tabular}{|c|c|c|c|c|c|c|}
% \hline
%  $Type\ Name$ & UCHAR & SCHAR & USHORTINT & SSHORTINT  & BYTE & BV16 \\\hline
%  $Range$ & 0..255 & -128..127 & 0..65.535  & -32.768..32.767  & -- & --\\ \hline
%  $Physical\ Size $ & 1 byte & 1 byte & 2 bytes & 2 bytes &  1 bytes & 2 bytes \\ \hline
% \end{tabular}
% \end{center}

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
 $Type\ Name$ & $Range$ & $Physical\ Size $\\\hline
 UCHAR & 0..255 & 1 byte\\\hline
 SCHAR & -128..127 & 1 byte  \\\hline
 USHORTINT & 0..65.535 & 2 byte \\\hline
 SSHORTINT & -32.768..32.767 & 2 byte \\\hline
 BYTE & -- & 1 bytes  \\\hline
 BV16 & -- & 2 bytes \\ \hline
\end{tabular}
\end{center}

\end{table}

@@

% We also created the following lemmas:
%
% $
% \begin{array}{l}
%  \forall (val) . (val \in \mathit{UCHAR} |
%  \mathit{byte\_uchar}(\mathit{uchar\_byte}(val)) = val) \land\\
%  \forall (by) . (by \in \mathit{BYTE} |
%  \mathit{uchar\_byte}(\mathit{byte\_uchar}(by)) = by)
% \end{array}
% $
%[NEW PARAGRAPH-TO-REVIEW]
%Similarly, several other general functions and lemmas were created for all other data types.
%Some times, we need create arithmetic and logic functions that are more specific for a
%determined computer architecture. These functions are developed in the module ALU 
%(arithmetic  logic unit). The following shows some examples of specific functions used in
%microcontroller model of Z80. For example, the function $\mathit{instruction_next}$
%receives a $\mathit{BYTE}$ and returns a $\mathit{BYTE}$ that contains a sequence of bit:
%a signal bit,

\section{A B model of the Z80 instruction set}
\label{sec:z80}

The \textit{Z80} is a CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports 158
different instructions, including all the 78 from Intel 8080 microprocessor, and all of them were
specified. These instructions are classified into these categories: load and exchange; block transfer and
search; arithmetic and logical; rotate and shift; bit manipulation; jump, call and return; input/output;
and basic cpu control. Each category of instruction has different elements of
specification.

This section shows the elements that make up the different types of instructions.
The main elements are specified in the microcontroller module Z80 and parts of it
are presented in this section~\ref{sec:z80}. Groups of registers are represented
by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables in the clause is
in the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined by the
clause \textit{OPERATIONS}, and the most of its specification constructions are
similar to constructions from traditional programming languages. Several
functions were created to Z80. General functions that can be used in others
microcontroller models are defined in the modules of data types. Specific
functions of the microcontroller are defined in the ULA (arithmetic logic unit).

@@

\subsection{Modelling registers and input and output ports} [TODO:I need review
the english]

\hspace*{0.10in}\it i\_  $\in$  \it BYTE  $\land$  \it r\_ $\in$  \it BYTE  $\land$

\hspace*{0.10in}\it im $\in$ (\it BIT $\times$ \it BIT\rm )  $\land$

% Features of microcontroller
 The internal registers contain 176 bits of reading/writing memory that are represented by
identifiers used as parameters in the instructions . It includes two sets of six general purpose
registers which may be used individually as 8-bits registers or as 16-bits register pairs.  The working registers
are represented by variable $\mathit{rgs8}$. The domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set
formed by identifiers of registers of 8 bits. These registers can be accessed in pairs, forming 16-bits,
resulting in another set of identifiers of 16-bits registers, named $\mathit{id\_reg16}$.

@@

\subsection{Flag register} [TODO:I need review the text]
Another important element is the ``z'' register ($\mathit{rgs8(z0)}$),  that is used as a flag register. This
register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning:
\begin{description}
  \item[$\mathit{bv\_get(rgs8(z0),0)}$] - The Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),1)}$] - The Add/Subtract bit
  \item[$\mathit{bv\_get(rgs8(z0),2)}$] - The Parity or Overflow bit
  \item[$\mathit{bv\_get(rgs8(z0),4)}$] - The Half Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),6)}$] - The Zero bit
  \item[$\mathit{bv\_get(rgs8(z0),7)}$] - The Sign bit 
\end{description}

These bits also can be used to specify security properties for microcontroller programs.

\textbf{Assuring the absence of overflow:}
 \emph{To assure that not to happen overflow, the developer can add this
 expression ($\mathit{bv\_get(rgs8(z0),0)} \neq 1 \land \mathit{bv\_get(rgs8(z0),2) \neq 1}$)
 in the invariant. By default, the overflow can happen, but many times it can be dangerous. 
 Then, the developer may prohibit its use. This restriction can also become the more difficult
 to specify.}

\subsection{Manipulation data functions from Z80}
[I need to review the text]\\ There are some specific functions
from Z80 to manipulate the data. In addressing mode, the function $\mathit{bv\_ireg\_plus\_d}$ is  used to indexed address. It receives the
value of register ($\mathit{ix}$ or $\mathit{iy}$) and the displacement to return the sum, the result is the dislocated address memory, see
its the definition.

@@

Another derived function is  $\mathit{bv\_(ireg\_plus\_d)}$, this returns the value in the memory  address
returned by $\mathit{bv\_ireg\_plus\_d}$ function and its definition is similar.

@@

\subsection{Program, stack and data memory}[I need review the text]

The Z80 uses a unique memory for storing program instructions, data stack and data work. The memory has 16-bits
addressing and each address represents a byte. Thus the data from the memory module is very simple, as shown below,
but an additional care must be taken to preserve the consistency of memory.

@@

In general, the instructions can access all memory address, but it is very dangerous. For add more security, is
important that the program instructions has the access limited by region. Thus the designer can specify address
regions to restrict the access from instructions. The address regions can be specified how shown below.
The constants ($\mathit{PRORGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$) defines respectively a range of restricted address
for: programs instructions, data work, data stack.

$
\begin{array}{l}
\mathit{PROGRAM\_R\_ADR} = 0..16384 \land\\
\mathit{DATA\_R\_ADR} = 16385..49151 \land\\
\mathit{STACK\_R\_ADR} = 49152..65535
\end{array}
$

\textbf{Assuring the absence of overlapping of address regions:}
 \emph{To assure that address regions are well defined, then the designer must to verify the below expression.}

@@

\textbf{Preserving the consistency of the memory:} \emph{In general, the access to
some address regions is dangerous. Then, in each instruction has a specific pre-condition that verify
if the new address memory, that will be updated, is member of its region. For example, the $\mathit{PUSH}$
program instruction allows write only in the region of stack ($\mathit{STACK\_R\_ADR}$).}

@@

There are many functions defined in the module \textit{ALU}. In general, these functions take basic definitions or functions
to build new more specific functions. The function $\mathit{half8UCHAR}$ is used to get the half part of $\mathit{UCHAR}$ value.
It is important to know the half carry and it is used in the function $\mathit{add8UCHAR}$.

@@

A related function to subtract operation is $\mathit{substract8UCHAR}$. There are the same functions for the
$\mathit{SCHAR}$ type, they are respectively $\mathit{add8SCHAR}$ and $\mathit{substract8SCHAR}$, all these
functions are of 8 bits ($\mathit{BYTE}$) and defined similarly. In the same way, the arithmetic functions for 16
bits ($\mathit{BV16}$) are defined.
The module ALU has several others functions, but only some simplest functions also are typed and explained below:

\begin{itemize}
  %\item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and
  %returns its increment. There is a similar function named $\mathit{dec}$.

  \item \it instruction\_next  $\in$  USHORT  $\fun$  USHORT \rm - It receives the  
  actual value from program counter register (\textit{pc}) and returns its increment.
 \item \it is\_negative  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns the most significant bit,
 in others words, the signal bit.

 % \item \it is\_zero  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
 % byte is zero, otherwise returns 0.

  %\item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.

@@

Every logic functions that are defined in the \textit{BYTE} and \textit{BV16} module are included in the \textit{ALU}
module. Then, these functions can be seen directly in the \textit{ALU} and \textit{Z80} module.

\subsection{Modelling the instructions}[I need review the text]

Each instruction is represented by a B operation in the module Z80. The main module (\textit{Z80}) has two
types of operations, a type represents the external actions and the other represents the instructions of
microcontrollers. The external actions are shown in the section ~\ref{sec:externalactions}.  A simple
example from instruction is a $\mathit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The tools B does not allow use parentheses in identifiers,
then the characters \{``('',``)''\} are replaced respectively by \{``9'',``0''\} in the real specification.} shown below.
Many times to model a instruction is necessary to use the pre-defined functions, this facilitates the construction of model. By default, all
parameters from operations are either predefined elements in the model or integers values in the decimal
representation. This instruction use the $\mathit{updateAddressMem}$ function from \textit{Memory} module
and it receives a address memory and its new memory value. After, it increments the program counter
($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).

@@

\subsection{Modeling the input/output instructions}

@@

The $\mathit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the follow B operation. It is from I/O group, then it receives a identifier of register
$\mathit{r}$ and, in this place, it stores the value of $\mathit{C}$ port address. Besides, it increments the
program counter and updates the flag registers.

@@

\subsection{Modeling the external actions}
\label{sec:externalactions}
